---
title: "Daily Preliminary Ridership Report"
author: "Samuel B. Carter"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(format(Sys.Date(), "%Y%m%d"), " Preliminary Ridership Report",'.html')) })
output:
  html_document:
    # toc: true
    df_print: paged
    fig_cap: true
    code_folding: hide
    
---

These data are for general ridership monitoring purposes only. These data are unofficial and subject to change at any time.

Data are sourced from fareboxes for local routes and automatic passenger counters for Routes 90, 901, and 902. 

```{r include = FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

```{r revision history}

# 5/7/20 - Created script.

```

```{r future improvements}

# 5/7/20 - Add comparison the previous day's results.
#           - one way may be to export csv, and then read-in that csv, comparing the two dates...
#        - Add good way to add 901/902.
#             - Alternatively, consider APCs for 901/902?
#        - Add error term 
#             - (ie, if bogus data, don't send report!)

# 5/13/20 - Consider leading with graph of last thirty days.
#         - Consider arranging table by most recent date.
#         - Consider function to search for *most recent* GFI export, in
#           case a day is missed... (UPDATE: DONE!)

# 5/15/20 - Consider adding simple narrative description of ridership change...
#         - Need to be able to upload data to sandbox..

# 5/17/20 - Another problem: because it looks for the last date, it won't work
#           *after* you run a report for the first time each day..

# 5/19/20 - I don't think the note from 5/17 is an issue.
#           However, should probably move operative folder to shared drive..



```

```{r notes}

# 5/7/20 - This does not currently include route 999. This means historical
#          data will be less accurate but more recent data will be more
#          accurate (I believe... or at least it'll compare to the daily report
#          better! Once those 999 routes are inputted.. then.. it won't match
#          the previous, daily report, but the totals should align with the revised report...
#          which leads me to believe they should be included!)

#          Update: let's include 999 for now. (It may also be worth researching whether 999 has increased
#          since the introduction of 901/902...)

```

```{r set libraries, include=FALSE, results='hide'}

library(tidyverse)
library(kableExtra)
library(scales)
library(lubridate)
library(odbc)

```

```{r define date variables}

# set epoch date variable

GFI_COVID_start <- as.Date("2020-03-01")

# set dynamic date variables 

GFI_yesterday <- lubridate::floor_date(Sys.Date() -1, unit = "day")

GFI_today <- lubridate::floor_date(Sys.Date(), unit = "day")

# set other variables

route_set <- c(2:6,8,10:16,18:19,21,24:26,
                      28,30,31,34,37:39,55,86,87, 999)
```

This report covers service between `r GFI_COVID_start` and `r GFI_yesterday`.

```{r get and clean data, results='asis'}
# gather, clean, transform data

# set GFI connection

con <- dbConnect(odbc(), "gfi_ds1_indy", Driver = "SQL Anywhere 12")
  
# get ev data

ev_recent_summary <- tbl(con, dbplyr::in_schema("DBA", "ev")) %>%
  filter(ts >= GFI_COVID_start, ts < GFI_today,
         route %in% route_set,
         rdr_c > 0) %>%
  mutate(Date = as.Date(str_sub(ts, 1, 10))) %>%
  group_by(Date) %>%
  summarise(rdr_sum = sum(rdr_c, na.rm = TRUE)) %>%
  arrange(Date) %>%
  collect()

# get APC data

con <- DBI::dbConnect(odbc::odbc(), Driver = "SQL Server", Server = "REPSQLP01VW", 
                      Database = "Steve_Temp", Port = 1433)

Apc_recent <- tbl(con, "Apc_Data") %>%
  filter(MDCID >= 1950, MDCID < 2000 | MDCID == 1899,
         GPS_Time >= "2020-02-29 00:00:00", 
         Route %in% c("90", "901", "902")) %>%
  collect()

```

```{r clean APC data, results='asis'}

Apc_recent$Date <- as.Date(str_sub(Apc_recent$GPS_Time, 1, 10))

Apc_recent$Clock_Time <- str_sub(Apc_recent$GPS_Time, 12, 19)

Apc_recent$DateTest <- ifelse(Apc_recent$Clock_Time < 
                                "03:00:00", 1, 0)

Apc_recent$Transit_Day_Unix <- ifelse(Apc_recent$DateTest == 1,
                                      lubridate::as_date(Apc_recent$Date - 1),
                                      Apc_recent$Date)

Apc_recent$Epoch_Date <- as.Date("1970-01-01")

Apc_recent$Transit_Day <- Apc_recent$Epoch_Date + lubridate::days(Apc_recent$Transit_Day_Unix)

```

```{r transform APC data, results='asis'}

# inspect ridership per trip per day...

APC_summary_90 <- Apc_recent %>%
  filter(Route %in% c("90")) %>%
  group_by(Transit_Day) %>%
  summarise(Boardings = sum(Boarding),
            Alightings = sum(Alighting),
            Difference = sum(Boarding) - sum(Alighting)) %>%
  arrange(desc(Transit_Day))

APC_summary_901 <- Apc_recent %>%
  filter(Route %in% c("901")) %>%
  group_by(Transit_Day) %>%
  summarise(Boardings = sum(Boarding),
            Alightings = sum(Alighting),
            Difference = sum(Boarding) - sum(Alighting)) %>%
  arrange(desc(Transit_Day))

APC_summary_902 <- Apc_recent %>%
  filter(Route %in% c("902")) %>%
  group_by(Transit_Day) %>%
  summarise(Boardings = sum(Boarding),
            Alightings = sum(Alighting),
            Difference = sum(Boarding) - sum(Alighting)) %>%
  arrange(desc(Transit_Day))

```

```{r merge 901/902 with ev_recent and clean, results = 'asis'}

ev_recent_summary_joined <- ev_recent_summary %>%
  left_join(APC_summary_901, by = c("Date" = "Transit_Day")) %>%
  left_join(APC_summary_902, by = c("Date" = "Transit_Day")) %>%
  mutate(Total_Boardings = rdr_sum + Boardings.x + Boardings.y)

ev_recent_summary_joined <- select(ev_recent_summary_joined, 
                                   Date, 
                                   Ridership = Total_Boardings)

```

```{r export GFI summary import old GFI summary, results = 'asis'}

write.csv(ev_recent_summary_joined, paste0(GFI_today, "-GFI-ev",".csv"), row.names = FALSE)

most_recent_GFI <- sort(str_sub(list.files()[str_which(list.files(), "-GFI-ev.csv$")], 1, 10), TRUE)[2]

yesterday_GFI <- read.csv(paste0(most_recent_GFI, "-GFI-ev",".csv"),
                          colClasses = c("Date" = "Date"))

```

```{r merge today and yesteday, make comparison, results = 'asis'}

ev_recent_summary_joined <- ev_recent_summary_joined %>%
  left_join(yesterday_GFI, by = "Date") %>%
  mutate('Ridership Change' = Ridership.x -  Ridership.y)

```

```{r merge Red Line, format table, results = 'asis'}

ev_recent_summary_joined_final <- ev_recent_summary_joined %>%
  left_join(APC_summary_90, by = c("Date" = "Transit_Day")) %>%
  mutate(System_Ridership = Boardings + Ridership.x ) %>%
  select(Date, 'Local Passenger Trips' = Ridership.x, 
         'Red Line Passenger Trips' = Boardings,
          'Fixed-Route Passenger Trips' = System_Ridership)

```

```{r make ggplot of past thirty days, results = 'asis', out.width = "100%", out.height= "100%"}

ev_recent_summary_joined_final %>% 
  filter(Date >= as.Date(lubridate::today(),format='%d-%B-%Y')-31) %>% # for past thirty days data
  mutate(Date = format(Date, '%m/%d   %a')) %>%
  mutate(Date = str_replace_all(Date, "^0[:digit:]/", str_sub(Date, 2, 3) )) %>%
  pivot_longer(cols = c('Local Passenger Trips', 'Red Line Passenger Trips'), names_to = "Mode") %>%
  ggplot() + 
  geom_bar(aes(x = factor(Date), y= value, 
               fill = factor(Mode, levels = c('Local Passenger Trips', 'Red Line Passenger Trips'))),
           stat = "identity", alpha = .8,
           position = "stack") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.1))+ # make size 6, if doing 60 day comparison
  labs(title = "Unofficial Fixed-Route Passenger Trips",
       subtitle = "Past Thirty Days",
       caption = "Figures are subject to change as data become more complete",
       y = "Boardings",
       x = "") + 
  scale_fill_discrete(name = "Mode") +
  ylim(0,40000)

```

### Unofficial Fixed-Route Passenger Trips
```{r Unofficial Fixed-Route Passenger Trips, results='asis' }


ev_recent_summary_joined_final %>%
  arrange(desc(Date)) %>%
  mutate('Day of Week' = lubridate::wday(Date, abbr = FALSE, label = TRUE),
         Date = format(Date, '%m/%d')) %>%
  select(Date, 'Day of Week', everything()) %>%
kable(., format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  full_width = F)

  
```

### Change since last Report
```{r Change since last report, results='asis'}

ev_recent_summary_joined_final %>%
  left_join(yesterday_GFI, by = "Date") %>%
  mutate('Day of Week' = lubridate::wday(Date, abbr = FALSE, label = TRUE),
         Date = format(Date, '%m/%d/%y')) %>%
  select(Ridership_today = 'Local Passenger Trips', everything(),
         RL = 'Red Line Passenger Trips') %>%
  mutate('Difference' = Ridership_today - Ridership, 
         'Ridership in Last Report' = RL + Ridership) %>%
  select(Date, 'Day of Week', 'Ridership Today' = 'Fixed-Route Passenger Trips',
         'Ridership in Last Report', 'Difference') %>%
  filter(Difference > 0) %>%
  arrange(desc(Date)) %>%
kable(., format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),  
                full_width = F,
                )

  
```

```{r ggplot since 3/2, results = 'asis', out.width = "100%", out.height= "100%"}

ev_recent_summary_joined_final %>% 
  filter(Date >="2020-03-02") %>% # since 3/2
  # mutate(Date = format(Date, '%m/%d   %a')) %>%
  # mutate(Date = str_replace_all(Date, "^0[:digit:]/", str_sub(Date, 2, 3) )) %>%
  pivot_longer(cols = c('Local Passenger Trips', 'Red Line Passenger Trips'), names_to = "Mode") %>%
  ggplot() + 
  geom_bar(aes(x = Date, y= value, 
               fill = factor(Mode, levels = c('Local Passenger Trips', 'Red Line Passenger Trips'))),
           stat = "identity", alpha = .8,
           position = "stack") +
  theme(axis.text.x = element_text(angle = 90, vjust=0.1))+ # make size 6, if doing 60 day comparison
  labs(title = "Unofficial Fixed-Route Passenger Trips",
       subtitle = "Since March 2nd, 2020",
       caption = "Figures are subject to change as data become more complete",
       y = "Boardings",
       x = "") + 
  scale_fill_discrete(name = "Mode") +
  ylim(0,40000) #+
  # scale_x_date(name = 'My date axis title', date_breaks = '2 days',
  #       date_labels = '%m/%d   %a',
  #       limits = as.Date(c("2020-03-01", GFI_today)),
  #       expand_scale(mult = 1, add = 0)) #+
  # xlim(as.Date(c('2020-03-02')), as.Date(GFI_yesterday))# come back to this...

```
If you believe this report is in error, please contact Samuel Carter (scarter@indygo.net) or Michael Nugent (michael.nugent@indygo.net). Report last revised 2020-08-06.